---
title: "Correlatie en Regressie Microcursus"
subtitle: "Statistiek in de Criminologie - Dodona Platform"
author: "Universiteit Gent - Criminologie Departement"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: hide
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(knitr)
library(DT)
```

# Correlatie en Regressie voor Criminologie

**Georganiseerd volgens Bloom's Taxonomie:**
- **Remember** (Herinneren): Basisbegrippen van correlatie en regressie oproepen
- **Understand** (Begrijpen): Concepten van verbanden en voorspellingen begrijpen  
- **Apply** (Toepassen): Correlatie- en regressietechnieken toepassen
- **Analyze** (Analyseren): Verbanden analyseren en interpreteren
- **Evaluate** (Evalueren): Kwaliteit van analyses kritisch beoordelen
- **Create** (Creëren): Eigen analyses en interpretaties ontwikkelen

Deze **angst-vriendelijke cursus** bestaat uit **4 modules** met **72 micro-vragen**. Elke oefening bevat eenvoudige uitleg, stapsgewijze hints en feedback bij veelgemaakte fouten.

**Totaal Vragen:** 72 micro-vragen verdeeld over 4 modules
**Student-vriendelijk Design:** Eerst visueel, dan getallen, één stap per scherm
**Externe Bronnen:** Betrouwbare educatieve links voor visueel leren

---

# Module 1: Correlation (Part 1: Basics & Intuition)
*18 micro-questions*

**Student Learning Outcomes:**
- Read scatterplots effectively
- Identify direction (↑/↓) and strength (weak/moderate/strong)
- Understand why covariance ≠ comparable
- Grasp unit-free correlation concept
- Master z-scores and (0,0) center
- Know correlation ≠ causation

## Question 1: Spot the Trend I
**Type:** Visual MCQ  
**Task:** Pick the best label for the given scatterplot

**Options:**
- A) Strong positive
- B) Moderate positive  
- C) Strong negative
- D) Moderate negative
- E) No clear pattern

**Learning Focus:** Basic pattern recognition in scatterplots

---

## Question 2: Spot the Trend II (Harder Clouds)
**Type:** Visual MCQ  
**Task:** Same as Q1 but with noisier, more ambiguous patterns

**Options:**
- A) Strong positive
- B) Moderate positive
- C) Weak positive
- D) Weak negative
- E) No linear relationship

**Learning Focus:** Pattern recognition with increased noise/ambiguity

---

## Question 3: Quadrants = Sign
**Type:** Visual MCQ  
**Task:** Points mainly in quadrants I & III vs II & IV → determine association sign

**Options:**
- A) Positive association (most points in I & III)
- B) Negative association (most points in II & IV)
- C) No clear pattern

**Learning Focus:** Understanding quadrant-based correlation signs

---

## Question 4: What Does z=+1.2 Mean?
**Type:** Short Answer  
**Task:** Explain "z=+1.2 on nuisance variable" in plain words

**Expected Answer:** "The value is 1.2 standard deviations above the mean"

**Learning Focus:** Z-score interpretation in context

---

## Question 5: Why Standardize? (Story)
**Type:** Concept MCQ  
**Task:** Income per year vs per week: does the relationship change? Why can't we compare raw covariance?

**Options:**
- A) Relationship changes because time units differ
- B) Relationship stays the same, but covariance values aren't comparable due to different units
- C) Covariance is always the better measure
- D) We should never standardize variables

**Learning Focus:** Understanding the need for standardization

---

## Question 6: Covariance vs Correlation (Drag-and-Drop)
**Type:** Matching Exercise  
**Task:** Match terms to their correct meanings

**Left Column:**
- Covariance
- Correlation coefficient

**Right Column:**
- Has units from original variables
- Always between -1 and +1
- Unit-free measure
- Can have unlimited range

**Learning Focus:** Distinguishing between covariance and correlation properties

---

## Question 7: Find the Center
**Type:** Visual MCQ  
**Task:** On a z-score plot, identify the bivariate "center of gravity"

**Options:**
- A) (0, 0) - the intersection of mean lines
- B) (1, 1) - one standard deviation point
- C) The highest point in the distribution
- D) The point with most data

**Learning Focus:** Understanding standardized coordinate systems

---

## Question 8: Direction & Strength from a Line
**Type:** Visual MCQ  
**Task:** Given trend line + scattered points, identify sign and strength

**Options:**
- A) Strong positive
- B) Moderate positive
- C) Weak positive
- D) Strong negative
- E) Moderate negative

**Learning Focus:** Interpreting fitted lines with actual data

---

## Question 9: Outlier Intuition
**Type:** Visual MCQ  
**Task:** Which outlier position will increase vs decrease correlation?

**Scenario:** Scatterplot with positive trend
**Options:**
- A) Top-right outlier will increase r
- B) Bottom-right outlier will decrease r
- C) Both A and B are correct
- D) Outliers don't affect correlation

**Learning Focus:** Understanding outlier impact on correlation

---

## Question 10: Same Stats, Different Shapes
**Type:** Concept MCQ  
**Task:** Four plots with identical correlation values but different patterns

**Question:** What's the main lesson from Anscombe's Quartet?
**Options:**
- A) Correlation tells the whole story
- B) Always plot your data first
- C) Statistics are more reliable than graphs
- D) Linear relationships are most common

**External Resource:** [Anscombe's Quartet - Wikipedia](https://en.wikipedia.org/wiki/Anscombe%27s_quartet)

---

## Question 11: Correlation ≠ Causation (Headline)
**Type:** Short Answer  
**Task:** Offer a plausible third variable (confounder) that could explain both variables

**Scenario:** "Ice cream sales and drowning incidents show strong positive correlation"
**Expected Answer:** Temperature, season, number of people at beaches/pools, etc.

**External Resource:** [Khan Academy: Correlation vs Causation](https://www.khanacademy.org/math/probability/xa88397b6%3Ascatterplots/estimating-trend-lines/v/correlation-and-causality)

---

## Question 12: Positive but Weak
**Type:** Visual MCQ  
**Task:** Distinguish between weak positive and no linear pattern

**Options:**
- A) Weak positive relationship
- B) No linear relationship
- C) Strong positive relationship
- D) Cannot determine from the plot

**Learning Focus:** Fine distinctions in weak associations

---

## Question 13: Monotone vs Linear
**Type:** Visual MCQ  
**Task:** Curved but monotone relation → choose appropriate correlation measure

**Scenario:** Data shows consistent upward trend but curved pattern
**Options:**
- A) Pearson correlation (measures linear relationship)
- B) Spearman correlation (measures monotonic relationship)
- C) Both are equally appropriate
- D) Neither is appropriate

**External Resource:** [Spearman's Rank Correlation - Laerd Statistics](https://statistics.laerd.com/statistical-guides/spearmans-rank-order-correlation-statistical-guide.php)

---

## Question 14: Units Change, r Doesn't
**Type:** Concept MCQ  
**Task:** If X is measured in cm instead of meters, what happens to correlation?

**Options:**
- A) Correlation increases by factor of 100
- B) Correlation decreases by factor of 100
- C) Correlation stays exactly the same
- D) Correlation becomes negative

**Learning Focus:** Unit invariance of correlation

---

## Question 15: Z-scores Help Compare
**Type:** Short Answer  
**Task:** Why do z-scores make axes comparable in a scatterplot?

**Expected Answer:** "Z-scores put both variables on the same scale (standard deviation units), making it easier to see the relationship without being distracted by different measurement units"

---

## Question 16: Reading a Legend
**Type:** Visual MCQ  
**Task:** Interpret color-coded subgroups correctly

**Scenario:** Scatterplot with points colored by neighborhood type
**Question:** The overall correlation is calculated using:
**Options:**
- A) Only the largest subgroup
- B) Each subgroup separately
- C) All points together, ignoring subgroup colors
- D) The average of subgroup correlations

**Learning Focus:** Understanding aggregate vs. subgroup analysis

---

## Question 17: Practice Labeling
**Type:** Short Answer  
**Task:** Describe the scatterplot relationship in one sentence using "as X increases..."

**Expected Format:** "As X increases, Y tends to [increase/decrease] [weakly/moderately/strongly]"

---

## Question 18: Mini-Reflection
**Type:** Free Response (2 lines)  
**Task:** Metacognitive reflection

**Prompts:**
- One thing I learned in this module:
- One thing that's still fuzzy:

**Learning Focus:** Anxiety reduction and metacognition

---

# Module 2: Correlation (Part 2: Do the Analysis)
*18 micro-questions*

**Student Learning Outcomes:**
- Compute mean, deviations, SSx, SSy, cross-products step-by-step
- Calculate covariance and Pearson r manually
- Interpret r and R² values
- Choose appropriate association measures for different data types

*Uses small datasets with 6-10 rows; auto-checks with ±0.01 tolerance*

## Question 19: Compute Means (X̄, Ȳ)
**Type:** Numeric Input  
**Data:** Sample dataset with Test1 and Test2 scores (8 rows)

```{r sample-data, echo=TRUE}
# Sample Data for Questions 19-26
test_data <- data.frame(
  Student = 1:8,
  Test1 = c(72, 85, 90, 78, 82, 88, 75, 92),
  Test2 = c(74, 88, 92, 80, 85, 90, 77, 95)
)
kable(test_data, caption = "Sample Dataset: Test Scores")
```

**Task:** Calculate X̄ (mean of Test1) and Ȳ (mean of Test2)
**Expected Answers:** X̄ = 82.75, Ȳ = 85.125

---

## Question 20: Compute Deviations
**Type:** Table Fill  
**Task:** Complete the table with (X−X̄) and (Y−Ȳ) columns

**Learning Focus:** Understanding deviations from the mean

---

## Question 21: Squares & Cross-Products
**Type:** Table Fill  
**Task:** Add columns for (X−X̄)², (Y−Ȳ)², and (X−X̄)(Y−Ȳ)

**Learning Focus:** Building blocks of correlation calculation

---

## Question 22: Sum of Squares
**Type:** Numeric Input  
**Task:** Calculate SSx, SSy, and SSxy

**Formulas:**
- SSx = Σ(X−X̄)²
- SSy = Σ(Y−Ȳ)²  
- SSxy = Σ(X−X̄)(Y−Ȳ)

---

## Question 23: Sample Variances & Standard Deviations
**Type:** Numeric Input  
**Task:** Calculate sample variances and standard deviations

**Formulas:**
- Var(X) = SSx/(n−1)
- SD(X) = √Var(X)
- Same for Y

**Common Mistake Feedback:** "Remember to use n−1 for sample variance!"

---

## Question 24: Covariance
**Type:** Numeric Input  
**Task:** Calculate sample covariance

**Formula:** Cov = SSxy/(n−1)
**Common Mistake:** Dividing by n instead of n−1

---

## Question 25: Pearson r (Route A: Standardized Covariance)
**Type:** Numeric Input  
**Task:** Calculate correlation using r = Cov / (Sx·Sy)

**Learning Focus:** Understanding r as standardized covariance
**External Resource:** [Penn State STAT 200 - Correlation](https://online.stat.psu.edu/stat200/lesson/3/3.4/3.4.2)

---

## Question 26: Pearson r (Route B: Z-scores)
**Type:** Numeric Input  
**Task:** Show that r = average of zₓ·zᵧ (conceptual formula)

**Learning Focus:** Conceptual understanding of correlation
**External Resource:** [Penn State STAT 200 - Correlation Formula](https://online.stat.psu.edu/stat200/lesson/12/12.2)

---

## Question 27: Interpret r Magnitude
**Type:** MCQ  
**Task:** Classify correlation strength using your calculated r

**Options:**
- A) Very weak (0.00 to 0.19)
- B) Weak (0.20 to 0.39)
- C) Moderate (0.40 to 0.69)
- D) Strong (0.70 to 0.89)
- E) Very strong (0.90 to 1.00)

**Note:** These are guidelines, not absolute rules

---

## Question 28: Compute R²
**Type:** Numeric + Short Answer  
**Task:** Calculate R² and interpret in one sentence

**Formula:** R² = r²
**Interpretation Format:** "About X% of Y's variance is associated with X"

---

## Question 29: Spearman from Ranks
**Type:** Table Fill  
**Task:** Rank both variables (handling ties with average rank) → compute Spearman's ρ

**Method:** Apply Pearson formula to ranked data
**External Resource:** [Spearman's Rank Correlation - Wikipedia](https://en.wikipedia.org/wiki/Spearman%27s_rank_correlation_coefficient)

---

## Question 30: Spearman: Ties Note (Concept)
**Type:** MCQ  
**Task:** Understanding why we adjust for ties

**Question:** Why do we use average ranks for tied values?
**Options:**
- A) It's easier to calculate
- B) It ensures fair treatment of equal values
- C) It makes the correlation stronger
- D) It's required by software

**Advanced Note:** Kendall's τ-b handles ties more gracefully
**External Resource:** [Kendall Rank Correlation - Wikipedia](https://en.wikipedia.org/wiki/Kendall_rank_correlation_coefficient)

---

## Question 31: Cramér's V (3×3)
**Type:** Numeric Input  
**Task:** Calculate Cramér's V from contingency table

**Given:** 3×3 contingency table
**Formula:** V = √(χ²/(n·(min(r−1,c−1))))
**Steps:**
1. Calculate χ²
2. Apply Cramér's V formula

**External Resource:** [Cramér's V - Wikipedia](https://en.wikipedia.org/wiki/Cram%C3%A9r%27s_V)

---

## Question 32: Point-Biserial r
**Type:** Numeric Input  
**Task:** Calculate correlation between metric and binary variables

**Scenario:** Continuous test scores vs. pass/fail (0/1)
**Method:** Standard Pearson correlation with 0/1 coding
**External Resource:** [Point-Biserial Correlation - Laerd Statistics](https://statistics.laerd.com/spss-tutorials/point-biserial-correlation-using-spss-statistics.php)

---

## Question 33: Choose the Right Measure (6 Scenarios)
**Type:** MCQ  
**Task:** Match correlation type to data types

**Scenarios:**
1. Height vs. Weight → **Pearson**
2. Education rank vs. Income rank → **Spearman**
3. Gender vs. Income → **Point-biserial**
4. Neighborhood type vs. Crime type → **Cramér's V**
5. Curved but monotonic relationship → **Spearman**
6. Linear relationship with outliers → **Pearson** (with caution)

**External Resource:** [UCLA OARC - Choosing Statistical Tests](https://stats.oarc.ucla.edu/other/mult-pkg/whatstat/)

---

## Question 34: Outlier Sensitivity
**Type:** Visual MCQ  
**Task:** Predict how adding an extreme point changes r

**Scenario:** Shows before/after scatterplots
**Options:**
- A) r increases substantially
- B) r decreases substantially
- C) r changes only slightly
- D) r becomes negative

---

## Question 35: Plot-First Check
**Type:** Visual MCQ  
**Task:** Recognize when Spearman is better than Pearson

**Scenario:** Nonlinear but monotone relationship
**Correct Choice:** Spearman (captures monotonic trend better)
**External Resource:** [Spearman vs Pearson - Laerd Statistics](https://statistics.laerd.com/statistical-guides/spearmans-rank-order-correlation-statistical-guide.php)

---

## Question 36: Mini-Recap
**Type:** Short Answer  
**Task:** Explain why r is easier to compare than covariance

**Expected Answer:** "Correlation is unit-free and bounded between -1 and +1, making it easy to compare strength across different variable pairs, while covariance depends on the original units"

---

# Module 3: Regression (Part 1: The Idea)
*17 micro-questions*

**Student Learning Outcomes:**
- Understand asymmetry (predict Y from X)
- Interpret intercept and slope (sign & magnitude)
- Grasp what residuals represent
- Understand OLS as "smallest squared vertical errors"
- See R² as "how tight points are around the line"

## Question 37: From Cloud to Line
**Type:** Visual MCQ  
**Task:** What does the best-fit line minimize?

**Options:**
- A) Horizontal distances from points to line
- B) Vertical distances from points to line
- C) Squared vertical distances from points to line
- D) Total number of points above vs. below line

**Correct Answer:** C (OLS principle)

---

## Question 38: Line Passes Through (X̄,Ȳ)
**Type:** True/False with Visual  
**Statement:** "The regression line always passes through the point (X̄,Ȳ)"

**Answer:** True
**Learning Focus:** Fundamental property of regression lines

---

## Question 39: Slope Sign
**Type:** Visual MCQ  
**Task:** Identify slope sign from line visualization

**Options:**
- A) Positive slope (line goes up-right)
- B) Negative slope (line goes down-right)
- C) Zero slope (horizontal line)

---

## Question 40: Slope Magnitude (Steeper = Bigger |b|)
**Type:** Visual MCQ  
**Task:** Compare slope magnitudes between two lines

**Question:** Which line has the larger absolute slope?
**Learning Focus:** Visual interpretation of slope steepness

---

## Question 41: Intercept Meaning
**Type:** Short Answer  
**Task:** Interpret intercept in context

**Prompts:**
- What does the y-intercept mean in this scenario?
- When might the intercept be meaningless or uninterpretable?

**Example Context:** Predicting crime rate from unemployment rate

---

## Question 42: Predicting Ŷ
**Type:** MCQ  
**Task:** Understanding slope interpretation

**Question:** If X increases by 1 unit, how much does Ŷ change?
**Options:**
- A) By 1 unit
- B) By the slope (b) units
- C) By the intercept (a) units
- D) It depends on the current X value

**Correct Answer:** B

---

## Question 43: Residuals on the Picture
**Type:** Visual MCQ  
**Task:** Identify residuals as vertical distances

**Visual:** Scatterplot with regression line and arrows showing residuals
**Task:** Distinguish over-prediction vs. under-prediction arrows

---

## Question 44: Why 'Squared' Errors?
**Type:** MCQ  
**Task:** Understand why OLS uses squared residuals

**Question:** Why do we square the residuals instead of using absolute values?
**Options:**
- A) Easier to calculate
- B) Penalizes large errors more heavily
- C) Prevents positive and negative errors from canceling out
- D) Both B and C are correct

**Correct Answer:** D

---

## Question 45: Two Fits, Same Slope
**Type:** Visual MCQ  
**Task:** Compare R² values for different scatter patterns

**Scenario:** Two scatterplots with identical slopes but different scatter
**Question:** Which has higher R²?
**Answer:** The one with points closer to the line

---

## Question 46: Standardizing X Helps Interpretation
**Type:** Concept MCQ  
**Task:** Understand effect of standardizing predictors

**Question:** When we z-score X, what changes in the regression interpretation?
**Options:**
- A) Slope now represents change per standard deviation of X
- B) Intercept becomes the predicted Y when X is at its mean
- C) Both A and B
- D) Nothing changes

**Correct Answer:** C

---

## Question 47: Unit Change Thought Experiment
**Type:** MCQ  
**Task:** Understand how rescaling affects regression components

**Scenario:** Measuring X in centimeters instead of meters
**Question:** What happens to the slope and correlation?
**Options:**
- A) Slope changes, correlation unchanged
- B) Both slope and correlation change
- C) Neither changes
- D) Only correlation changes

**Correct Answer:** A

---

## Question 48: Extrapolation Caution
**Type:** MCQ  
**Task:** Understand risks of predicting outside observed range

**Question:** Why is it risky to predict Y for X values far outside the observed range?
**Options:**
- A) The relationship might not be linear at extreme values
- B) We have no data to verify the pattern continues
- C) Confidence in predictions decreases
- D) All of the above

**Correct Answer:** D

---

## Question 49: Correlation vs Regression
**Type:** MCQ  
**Task:** Distinguish symmetric vs. asymmetric relationships

**Question:** What's the key difference between correlation and regression?
**Options:**
- A) Correlation measures strength, regression measures direction
- B) Correlation is symmetric, regression predicts Y from X
- C) Correlation uses all data, regression uses samples
- D) No meaningful difference

**Correct Answer:** B

---

## Question 50: Causality Caveat
**Type:** Short Answer  
**Task:** Explain why good fit ≠ causation

**Prompt:** Give one reason why a strong regression relationship doesn't prove causation
**Expected Answers:** Confounding variables, reverse causation, spurious correlation, etc.

---

## Question 51: Plot First (Again!)
**Type:** Visual MCQ  
**Task:** Recognize when linear regression is inappropriate

**Scenario:** Curved relationship fitted with straight line
**Question:** What should we consider?
**Options:**
- A) Linear regression is always appropriate
- B) Consider data transformation or nonlinear models
- C) Ignore the curvature
- D) Use correlation instead

**Correct Answer:** B

---

## Question 52: Noise and Uncertainty
**Type:** MCQ  
**Task:** Understand why points don't lie exactly on the line

**Question:** In social science data, why don't points lie exactly on the regression line?
**Options:**
- A) Measurement error
- B) Unobserved factors affecting Y
- C) Random variation
- D) All of the above

**Correct Answer:** D

---

## Question 53: Mini-Reflection
**Type:** Free Response (2 lines)  
**Task:** Metacognitive reflection on regression concepts

**Prompt:** One thing about regression that now feels simple to you

---

# Module 4: Regression (Part 2: Compute, Predict, Interpret)
*19 micro-questions*

**Student Learning Outcomes:**
- Compute b = Cov(X,Y)/Var(X) and a = Ȳ − bX̄
- Make predictions Ŷ = a + bX
- Understand R² as explained/total variance
- Connect standardized slope β = r·(Sy/Sx)
- Read basic statistical output

*Uses extended Test1-Test2 dataset with 10-15 rows*

## Question 54: Re-compute Means & SS
**Type:** Numeric Input  
**Task:** Quick recap of basic calculations

**Required:** X̄, Ȳ, SSx, SSy, SSxy
**Purpose:** Foundation for regression calculations

---

## Question 55: Compute Covariance & Var(X)
**Type:** Numeric Input  
**Task:** Calculate components needed for slope

**Formulas:**
- Cov = SSxy/(n−1)
- Var(X) = SSx/(n−1)

---

## Question 56: Slope b
**Type:** Numeric Input  
**Task:** Calculate regression slope

**Formula:** b = Cov(X,Y) / Var(X)
**Alternative:** b = SSxy / SSx

---

## Question 57: Intercept a
**Type:** Numeric Input  
**Task:** Calculate y-intercept

**Formula:** a = Ȳ − b·X̄
**Learning Focus:** Line passes through (X̄,Ȳ)

---

## Question 58: Predict Ŷ for Three X Values
**Type:** Numeric Input  
**Task:** Make predictions using Ŷ = a + bX

**Test Values:**
- X = X̄ (should give Ŷ = Ȳ)
- X = X̄ + 1SD
- X = X̄ − 1SD

**Check:** When X = X̄, Ŷ should equal Ȳ

---

## Question 59: Residuals for Three Cases
**Type:** Numeric Input  
**Task:** Calculate residuals e = Y − Ŷ

**Learning Focus:** Understanding prediction errors (sign and magnitude)

---

## Question 60: R² via r² (Bivariate)
**Type:** Numeric + Interpretation  
**Task:** Calculate R² and write interpretation

**Formula:** R² = r² (for simple regression)
**Interpretation:** "About ___% of the variance in Y is explained by X"

---

## Question 61: R² Intuition
**Type:** Visual MCQ  
**Task:** Compare R² values between plots

**Scenario:** Two scatterplots with regression lines
**Question:** Which has larger R²? Why?
**Learning Focus:** R² as measure of fit quality

---

## Question 62: Standardized Slope β
**Type:** Numeric Input  
**Task:** Calculate standardized regression coefficient

**Formula:** β = r·(Sy/Sx)
**Special Case:** If X and Y are z-scored, then β = r

---

## Question 63: Unit Changes
**Type:** MCQ  
**Task:** Understand what changes and what doesn't when rescaling

**Question:** If we rescale X, which remain unchanged?
**Options:**
- A) Raw slope (b) and intercept (a)
- B) Correlation (r) and R²
- C) Standardized slope (β)
- D) Both B and C

**Correct Answer:** D

---

## Question 64: Read SPSS-Style Output
**Type:** Pick-and-Label  
**Task:** Interpret regression output table

**Sample Output:**
```
                  B      Beta     t      Sig.
(Constant)     45.2              3.21   .003
Unemployment    2.8     .67     4.15   .001

R = .67, R² = .45
```

**Tasks:**
- Identify intercept: ___
- Identify slope: ___
- Identify standardized slope: ___
- Interpret R²: ___

---

## Question 65: Effect of an Outlier
**Type:** Visual MCQ  
**Task:** Predict outlier effects on regression

**Scenario:** Add extreme point to scatterplot
**Question:** How will this outlier affect slope and R²?
**Options:** Depends on outlier position relative to main trend

---

## Question 66: When is Intercept Meaningless?
**Type:** MCQ  
**Task:** Recognize interpretation limitations

**Scenario:** Predicting income from years of education
**Question:** When is the intercept (income at 0 years education) problematic?
**Answer:** When X=0 is outside realistic range or theoretically impossible

---

## Question 67: Check Model Assumptions (Lite)
**Type:** MCQ  
**Task:** Basic awareness of regression assumptions

**Key Assumptions:**
- Linearity
- Homoscedasticity (constant variance)
- Independence of observations

**Note:** We don't test these deeply—just awareness

---

## Question 68: Predictions vs Explanations
**Type:** Short Answer  
**Task:** Distinguish descriptive vs. causal interpretation

**Prompt:** In one sentence, what's the difference between predicting Y and explaining Y?

**Example Answer:** "Prediction focuses on accuracy of forecasting, while explanation implies understanding causal mechanisms"

---

## Question 69: Binary Y Caution
**Type:** MCQ  
**Task:** Recognize limitations of linear regression for binary outcomes

**Question:** Why shouldn't we use linear regression for predicting arrest (yes/no)?
**Options:**
- A) Predictions can fall outside 0-1 range
- B) Assumes linear probability change
- C) Violates normality assumptions
- D) All of the above

**Preview:** Logistic regression is more appropriate (not covered in this course)

---

## Question 70: From r to Slope
**Type:** Numeric Input  
**Task:** Convert correlation to regression slope

**Given:** r, Sx, Sy
**Formula:** b = r·(Sy/Sx)

---

## Question 71: From b Back to r
**Type:** Numeric Input  
**Task:** Convert regression slope to correlation

**Given:** b, Sx, Sy
**Formula:** r = b·(Sx/Sy)

---

## Question 72: Mini-Recap
**Type:** Short Answer  
**Task:** Final reflection on R²

**Prompt:** "Explain R² in your own words (not math formulas)"

**Example Answer:** "R² tells us what percentage of the variation in Y we can account for using X. Higher R² means the regression line fits the data better."

---

# Appendix: External Resources

## Interactive Learning Tools
- **Seeing Theory** - Interactive correlation & regression visualizations  
  [seeing-theory.brown.edu](https://seeing-theory.brown.edu/regression-analysis/index.html)

## Educational Videos
- **Khan Academy** - Correlation vs causation explanations  
  [Khan Academy Correlation](https://www.khanacademy.org/math/probability/xa88397b6%3Ascatterplots/estimating-trend-lines/v/correlation-and-causality)

## Statistical Concepts
- **Penn State STAT 200** - Correlation and regression theory  
  [Penn State Statistics](https://online.stat.psu.edu/stat200/lesson/12)
- **UCLA OARC** - Statistical test selection guide  
  [UCLA Statistical Tests](https://stats.oarc.ucla.edu/other/mult-pkg/whatstat/)

## Specialized Topics
- **Anscombe's Quartet** - Why visualization matters  
  [Wikipedia: Anscombe's Quartet](https://en.wikipedia.org/wiki/Anscombe%27s_quartet)
- **Spearman Correlation** - Rank-based correlation  
  [Spearman's Rank Correlation](https://en.wikipedia.org/wiki/Spearman%27s_rank_correlation_coefficient)
- **Cramér's V** - Effect size for categorical associations  
  [Cramér's V](https://en.wikipedia.org/wiki/Cram%C3%A9r%27s_V)

---

# Implementation Notes

## Anxiety-Friendly Design Principles
1. **Visual first, numbers second** - Start with plots, then calculations
2. **One tiny step per screen** - No overwhelming walls of text
3. **Progressive hints** - Reveal help gradually
4. **Gentle grading** - Tolerance for rounding, supportive feedback
5. **Metacognitive prompts** - Regular reflection opportunities

## Technical Specifications
- **Numeric tolerance:** ±0.01 for calculated values
- **Dataset sizes:** 6-15 rows for hand calculations
- **Progressive complexity:** Build from simple concepts to applications
- **External links:** Open in new tabs (Ctrl+click / Cmd+click)

## Assessment Features
- **Immediate feedback** - Right/wrong with explanations
- **Common mistake alerts** - Address frequent errors
- **Worked solutions** - Available after 2 attempts
- **Student Q&A** - Discussion tab for each question

This comprehensive question bank provides the foundation for implementing an engaging, supportive statistics course that builds student confidence while maintaining academic rigor.

---

# Module 5: Partiële correlatie & statistische controle
*7 micro-questions (Q73–Q79)*

**Student Learning Outcomes:**
- Begrijpen wanneer en waarom we partiële correlatie gebruiken  
- De formule voor \( r_{XY\cdot Z} \) toepassen  
- Schijnverband / suppressie herkennen  
- Resultaten kort en correct interpreteren in criminologische context

## Question 73: Zero-order vs. partiële correlatie (concept)
**Type:** MCQ  
**Task:** Wanneer is partiële correlatie gepast?

**Opties:**
- A) Als we willen weten of de relatie tussen X en Y blijft bestaan **na controle voor** Z  
- B) Als X en Y in verschillende meeteenheden staan  
- C) Als X en Y ordinaal zijn  
- D) Als de steekproef zeer klein is (n < 10)

**Correct:** A  
**Learning Focus:** Partiële correlatie = “netto”-verband tussen X en Y na statistische controle.

---

## Question 74: De formule
**Type:** Fill-in (formule)  
**Task:** Vul de standaardformule in voor partiële correlatie van X en Y gecontroleerd voor Z:

\[
r_{XY\cdot Z} = \frac{\,r_{XY} - r_{XZ}r_{YZ}\,}{\sqrt{(1-r^2_{XZ})(1-r^2_{YZ})}}
\]

**Learning Focus:** Notatie en structuur van de formule.

---

## Question 75: Rekenvraag met matrix (leeftijd als controle)
**Type:** Numeric Input (2 dec.)  
**Gegeven:** \( r_{XY}=0.97,\ r_{XZ}=0.80,\ r_{YZ}=0.85 \)  
**Task:** Bereken \( r_{XY\cdot Z} \) tot 2 decimalen.  
**Verwacht:** **0.92** (exact ≈ 0.918)  
**Interpretatie (korte zin):** De sterke relatie tussen numerieke en verbale scores blijft hoog na controle voor leeftijd.

---

## Question 76: Arrestaties vóór/na (criminologische casus)
**Type:** Numeric + Interpretatie (1 zin)  
**Gegeven:** \( r_{XY}=0.72 \) (Jaren gevangenisstraf & Arrestaties na), \( r_{XZ}=0.63 \) (Jaren gevangenisstraf & Arrestaties voor), \( r_{YZ}=0.76 \) (Arrestaties na & Arrestaties voor)  
**Task:** Bereken \( r_{XY\cdot Z} \) en interpreteer.  
**Verwacht:** **0.48** (exact ≈ 0.478)  
**Voorbeeldinterpretatie:** De relatie verzwakt maar blijft positief en matig na controle voor eerdere arrestaties—dus niet louter een artefact van het strafblad.

---

## Question 77: Schijnverband of suppressie?
**Type:** MCQ (scenario)  
**Scenario:** \( r_{XY}=0.18 \), \( r_{XY\cdot Z}=0.42 \) na controle voor Z.  
**Vraag:** Wat is de beste duiding?

**Opties:**
- A) Schijnverband (de relatie verdwijnt na controle)  
- B) Suppressie (de relatie wordt sterker na controle)  
- C) Geen verschil  
- D) Meetniveauprobleem

**Correct:** B (suppressie)  
**Learning Focus:** Netto-relatie kan sterker worden door Z te controleren.

---

## Question 78: Interactie ≠ controle
**Type:** Concept MCQ  
**Task:** Wat is geen juiste beschrijving van partiële correlatie?

**Opties:**
- A) Netto-samenhang tussen X en Y nadat Z constant is gehouden  
- B) Test of het effect van X op Y afhangt van de waarde van Z  
- C) Corrigeert voor mogelijke schijnverbanden  
- D) Gebruikt zero-order correlaties als input

**Correct:** B (dat beschrijft interactie/moderatie, niet partiële correlatie).

---

## Question 79: Korte interpretatie in context
**Type:** Short Answer (2 lijnen)  
**Prompt:** Je vindt \( r_{XY}=0.35 \) en \( r_{XY\cdot Z}=0.10 \). Schrijf in 1–2 zinnen wat dit betekent voor beleid/onderzoek (criminologische duiding).

---

# Add-on: Rapportagevaardigheden (uitbreiding bij regressie)
*2 micro-questions (Q80–Q81)*

**Doel:** Aansluiten bij de boeksectie over rapportage (B, β, R², SE, t, p) met criminologische duiding.

## Question 80: Schrijf de rapportagezin (APA-achtig)
**Type:** Write-the-sentence  
**Output (voorbeeld):**

```
                  B      Beta     SE      t      p
(Constant)     26.82            7.81    3.44   .004
Test_1          1.21    .80     0.25    4.78   <.001
R = .80, R² = .64
```

**Task:** Schrijf één correcte rapportagezin met effectschatting, significantie en duiding (geen causaliteit).  
**Voorbeeldsleutel (indicatief):**
“Een hogere score op Test_1 voorspelde hogere Test_2-scores (B=1.21, SE=0.25, β=.80, t=4.78, p<.001), met R²=.64; ca. 64% van de variantie in Test_2 werd verklaard.”

---

## Question 81: Duid intercept & schaal-invariantie
**Type:** MCQ (+ korte toelichting)  
**Vraag 1:** Wat gebeurt er met r en R² als X wordt omgerekend van meters naar centimeters?

**Opties:**
- A) r en R² veranderen mee  
- B) r en R² blijven gelijk  
- C) r neemt toe; R² daalt  
- D) Alleen r verandert

**Correct:** B (r en R² zijn schaal-invariant).  
**Vraag 2 (Short):** Wanneer is het intercept inhoudelijk minder zinvol?  
**Verwacht:** Wanneer X=0 buiten bereik of theoretisch onzinnig is.
